Why Donors Donate: Disentangling Organizational and Structural Heuristics for International Philanthropy

### 2025-01-11 manuscript/manuscript.qmd
Update updated citekeys

Non-governmental organizations (NGOs) face stricter regulations from their host governments [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have cracked down on these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. These attacks---most of which have been implemented through anti-NGO laws---have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official donors channel fewer funds to contentious issues such as human rights, elections, and advocacy, and instead choose to fund "tame" causes bush as health, education, and development, among others [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to individual donors abroad to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajorMello-e-Souza:2012; @Koch:2009; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organizational-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

It is possible to study the effects of both organizational and structural heuristics on donor decision making individually, but disentangling the interaction of the two heuristics from individual donor characteristics adds complex dimensionality and makes more standard experimental work difficult, costly, and statistically fraught. To address this, we use conjoint analysis to simultaneously test individual donor preferences in the face of both organizational and structural heuristics. Conjoint analysis is generally applicable whenever individuals face complex, multi-dimensional tradeoffs that make it difficult (or even impossible) to explicitly elicit preferences. Conjoint designs have been used extensively in political science to study complex preferences for elections, laws, public policy, and public opinion [@HainmuellerHopkinsYamamoto:2014; @Mummolo:2016; @KertzerRenshonYarhi-Milo:2021; @BansakHainmuellerHopkins:2021; @BremerBurgisser:2023], and conjoint experiments are becoming more common in nonprofit studies, particularly for studying individual donor behavior [@SilviaChildWitesman:2023; @HirschmannMoritzBlock:2022; @BachkeAlfnesWik:2014; @Kottasz:2004]. @KnudsenJohannesson:2019 argue that research related to respondent opinions requires conjoint analysis because it accounts for the masking effects of factors as well as the complexity of a respondent's judgment of credibility. Because we also want to capture the complexity of a respondent's judgment of philanthropic preferences, we use a conjoint experiment to measure the causal effect of both organizational and structural heuristics on individuals' preferences for donating to international NGOs.

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. Our reported causal effects are therefore not representative of the entire population and instead are only valid for people interested in charitable giving. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018]. Following screening, we received 1,016 viable responses. @apptbl-sample-details-demographics, @apptbl-sample-details-charity, and @apptbl-sample-details-politics provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity.

@@

[^stan]: We use Stan [@Stan-2-26-1 v2.26.1; @cmdstan-2-3-1 v2.3.1] through R [@rproject-4-3-1 v4.3.1] and **brms** [@Burkner:2017 v2.20.5] to estimate the model. We simulate 4 MCMC chains with 5,000 draws in each chain, 1,000 of which are used for warmup, resulting in 16,000 (4,000 × 4) draws per model parameter. We assess convergence with visual inspection, and all chains converge. Complete results from all the models, along with posterior predictive checks, goodness-of-fit measures, and model diagnostics---as well as our code and data---are available `ANONYMIZED_URL`.

@@

We provide viusalizations of the full posterior distributions of each of the effects of interest, and we report two distributional summary statistics: (1) the posterior median, (2) credible intervals based on a 95% equal-tailed quantile interval, However, we are generally less concerned with the exact point estimates of our causal effects and instead focus on the direction and relative magnitude of their posterior distributions. For inference, we calculate the probability of direction ($p_d$), or the probability that the posterior AMCE is strictly positive or negative---it is the proportion of the posterior AMCE that is the sign of the median.

>>>>

Non-governmental organizations (NGOs) face stricter regulations from their host governments [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @GlasiusSchalkDeLange:2020; @BakkeMitchellSmidt:2020]. Repressive governments have cracked down on these organizations using both violence, as well as administrative crackdown, which create barriers to funding and advocacy for a variety of NGOs [@DupuyRonPrakash:2016; @Chaudhry:2022; @Heiss:2017]. Policymakers have referred to this phenomenon as the "closing of civic space" or "shrinking civic space" [@CarothersBrechenmacher:2014; @Rutzen:2015b]. Data from the CIVICUS Civil Society Monitor shows that civil society groups face serious restrictions in more than 60% of countries---117 in total [@CIVICUSMonitor]. These attacks---most of which have been implemented through anti-NGO laws---have adversely affected the ability of both domestic and international NGOs to operate and raise funds. Official aid funds have decreased substantially in countries that have imposed restrictive anti-NGO legislation [@ChaudhryHeiss:2018; @DupuyPrakash:2018] and official donors channel fewer funds to contentious issues such as human rights, elections, and advocacy, and instead choose to fund "tame" causes bush as health, education, and development, among others [@ChaudhryHeiss:2018; @Bush:2015; @Herrold:2020; @RightSpringmanWibbels:2022]. In response to this dramatic shift in the NGO funding landscape, international NGOs (INGOs) have increasingly turned to individual donors abroad to offset the loss of funding from government and foundation sources [@BanksHulmeEdwards:2015].

@@

Our findings make two main contributions. First, existing research on privately-provided foreign aid primarily looks at organizations like foundations and large donor NGOs, and not private individuals [@ButheMajordeMelloeSouza:2012; @Koch:2009; @DesaiKharas:2018], and studies that do examine individual giving behavior typically only examine wealthy individuals [@USTrust2014]. Our research expands work on private international philanthropy by exploring how non-elite individuals decide to donate to NGOs focused on international affairs. Second, research shows that NGO leaders often have difficulty in attracting new donors and also find it challenging to motivate current donors [@GaudreauCao:2015]. Examining organizational-level factors that make the average donor increase their charitable giving, especially when facing restricted civic space abroad, can help NGOs better frame their funding appeals. Moreover, our results provide guidance to organizations facing increased antagonism by governments abroad and point to possible methods of adapting to worsening or shrinking civic space.

@@

It is possible to study the effects of both organizational and structural heuristics on donor decision making individually, but disentangling the interaction of the two heuristics from individual donor characteristics adds complex dimensionality and makes more standard experimental work difficult, costly, and statistically fraught. To address this, we use conjoint analysis to simultaneously test individual donor preferences in the face of both organizational and structural heuristics. Conjoint analysis is generally applicable whenever individuals face complex, multi-dimensional tradeoffs that make it difficult (or even impossible) to explicitly elicit preferences. Conjoint designs have been used extensively in political science to study complex preferences for elections, laws, public policy, and public opinion [@HainmuellerHopkinsYamamoto:2014; @Mummolo:2016; @KertzerRenshonYarhi-Milo:2021; @BansakHainmuellerHopkins:2021; @BremerBurgisser:2023], and conjoint experiments are becoming more common in nonprofit studies, particularly for studying individual donor behavior [@SilviaChildWitesman:2023; @HirschmannMoritzBlock:2022; @BachkeAlfnesWik:2014; @Kottasz:2004]. @KnudsenJohannesson:2018 argue that research related to respondent opinions requires conjoint analysis because it accounts for the masking effects of factors as well as the complexity of a respondent's judgment of credibility. Because we also want to capture the complexity of a respondent's judgment of philanthropic preferences, we use a conjoint experiment to measure the causal effect of both organizational and structural heuristics on individuals' preferences for donating to international NGOs.

@@

We fielded our survey experiment through Centiment, which recruits representative samples of paid (and highly engaged) survey participants online. To see how varying NGO characteristics influence the decision to donate, our sample was representative of the population of people who are likely to donate to charity. We asked potential participants a screening question about their philanthropic behavior early in the survey---if a participant responded that they give once every few years or never, they were disqualified from the study. Our reported causal effects are therefore not representative of the entire population and instead are only valid for people interested in charitable giving. We also included an attention check question early in the survey and removed respondents who failed the question. Importantly, these screening questions were presented *prior to* the experimental manipulation to avoid post-treatment bias [@MontgomeryNyhanTorres:2018a]. Following screening, we received 1,016 viable responses. @apptbl-sample-details-demographics, @apptbl-sample-details-charity, and @apptbl-sample-details-politics provide a detailed breakdown of the individual characteristics of our sample. In general, respondents were well balanced across all pre-treatment characteristics, including gender, age, education, income, and attitudes toward charity.

@@

[^stan]: We use Stan [@Stan-2-26-1 v2.26.1; @cmdstan-2-3-1 v2.3.1] through R [@rproject-4-3-3 v4.3.1] and **brms** [@Burkner:2017 v2.20.5] to estimate the model. We simulate 4 MCMC chains with 5,000 draws in each chain, 1,000 of which are used for warmup, resulting in 16,000 (4,000 × 4) draws per model parameter. We assess convergence with visual inspection, and all chains converge. Complete results from all the models, along with posterior predictive checks, goodness-of-fit measures, and model diagnostics---as well as our code and data---are available `ANONYMIZED_URL`.

@@

We provide visualizations of the full posterior distributions of each of the effects of interest, and we report two distributional summary statistics: (1) the posterior median, (2) credible intervals based on a 95% equal-tailed quantile interval, However, we are generally less concerned with the exact point estimates of our causal effects and instead focus on the direction and relative magnitude of their posterior distributions. For inference, we calculate the probability of direction ($p_d$), or the probability that the posterior AMCE is strictly positive or negative---it is the proportion of the posterior AMCE that is the sign of the median.
